---
title: "HW2 p8105 James Dalgleish jld2227"
author: "James Dalgleish"
date: "September 27, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
## Problem 1
Instructions(I'm choosing to include these to make it a bit easier to grade):
"Read and clean the data; retain line, station, name, station latitude / 
longitude, routes served, entry, vending, entrance type, and ADA compliance.
Convert the entry variable from character (YES vs NO) to a logical variable
(the ifelse or recode function may be useful)."
```{r data_import}
subway_data = read_csv("./problem1/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>% 
  select(c("line","station_name","station_latitude","station_longitude",
           starts_with("route"),"entry","vending","ada")) %>% 
  mutate(entry = recode(entry,"YES" = TRUE, "NO" = FALSE))
subway_data 
```

Instruction: "Write a short paragraph about this dataset – 
explain briefly what variables the dataset contains, 
describe your data cleaning steps so far,
and give the dimension (rows x columns) of the resulting dataset.
Are these data tidy?"

Essentially, the dataset contains two character variables dening the subway line and station,
two numeric (double) variables denoting geolocation (latitude and longitude),
11 variables denoting the routes (mostly populated with missing values) in a wide format,
two logical TRUE/FALSE variables for entry and ada,
and a character variable with values "YES" and "NO" variable for vending.
The data is not tidy for at least the reasons that routes are in a wide format 
where the route is encoded in the column. This can be remedied with the 
gather function. As we'll see below, route1 was been encoded in
mixed case(see the "E" and "e"?),
which may mess up some downstream analysis as well.

The dataset contains `r nrow(subway_data)` rows and `r ncol(subway_data)`
columns. Data cleaning included reading the data into a tibble, 
selecting the necessary columns, converting the column names 
into a standard format, and converting the entry variable into a logical column.
```{r describe}
skimr::skim(subway_data)
sapply(subway_data,table)
subway_data %>% 
  select(starts_with("route")) %>%
  sapply(.,table)
```
 Putting the wide format into long could aid analysis and can be done with a simple gather command.
```{r wide_to_long}
subway_data_long = subway_data %>% 
  gather(key = "route_number",value = "train",route1:route11)
```

Instruction: "Answer the following questions using these data:

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.
How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?"

We'll now select the columns of interest and limit to only the distinct rows,
then count the number of distinct rows containing route number and train
combinations.
```{r distinct_stations}
subway_routes_trains <- subway_data_long %>% select("route_number","train", "ada","entry") %>% distinct()
subway_routes_trains
nrow(subway_routes_trains)
```
There happen to be `nrow(subway_routes_trains)` distinct stations, according to the data.

# Ada compliance (recheck ths to make sure  that the problem is still correct).
We will create an additional variable-- station ada compliant. There can be multiple entrances, only some of which are ada compliant. A station is compliant if there is at least one station name/line combination with an ada compliant entrance, then it is ada compliant. Not all entrances need have an elevator.
# Without vending and allow entry.

Reformat data so that route number and route name are distinct variables. 
How many distinct stations serve the A train? How many are ADA compliant?
